---
title: "Data Science Term Project"
author: "Tamas Haller"
date: 'March 19th 2017 '
output: 
  html_document: 
    fig_caption: yes
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Subject
Almost everyone asked themselves at least once the following question: *What influences love at first site?*. In order to get some insights I downloaded a Speed dating data set from kaggle. 

The data set was created by Columbia Business School professors for their paper *Gender Differences in Mate Selection: Evidence From a Speed Dating Experiment*.

The data in the data set was collected from a participants of multiple experimental speed dating events. The speed dating experiment setup was the following:

* 4 minute conversation with a participant of opposite sex
* Complete a questionnaire about the "first date"
    + Would she/he like to see the date again?
    + Rate their date on six attributes: Attractiveness, Sincerity, Intelligence, Fun, Ambition, and Shared Interests
* Switch partners
* Continue the previous steps until all the participants met

From the above data set based on the provided variable I would like to predict if there would or would not be a match between two persons. There is a match is both persons answered that they would like to see each other again. Hopefully the analysis will also uncover what are the most desirable attributes in the opposite partner.  

## Data Source
I find the data under the following kaggle page: <https://www.kaggle.com/annavictoria/speed-dating-experiment>. The csv file itself could be downloaded from the following link: <http://www.stat.columbia.edu/~gelman/arm/examples/speed.dating/>
```{r speeddating, warning=FALSE, message=FALSE}
# LIBRARIES
library(arm)
library(readr)
library(dplyr)
library(ggplot2)
library(pastecs)
library(DataCombine)
library(descr)
library(stargazer)
library(sandwich)
library(lmtest)
library(splines)
library(readr)
library(mfx)
library(descr)
library(pander)
library(pROC)
library(data.table)
library(rpart)
library(partykit)
library(class)
library(caret)


# CLEAR MEMORY
rm(list = ls())

# CKECK  WORKING DIRECTORY
getwd()
setwd("C:\\Users\\IBM_ADMIN\\Desktop")

# LOAD  DATA
sdd <- read.csv("Speed Dating Data.csv")
```

## Data at first sight

After checking the data at first sight and reading it's description (which can be found at <http://www.stat.columbia.edu/~gelman/arm/examples/speed.dating/Speed%20Dating%20Data%20Key.doc>) we can see that it contains 8378 observations and 195 variables.

```{r check}
pander(colnames(sdd))
```

## Data cleaning

As a first step of data cleaning we will remove some of columns (variables) that are irrelevant from the perspective of our project and/or are not described by the documentation.

```{r datacleaning, echo=FALSE}
sdd <- sdd[,c(1:69)]

sdd$iid<-NULL
sdd$id<-NULL
sdd$idg<-NULL
sdd$condtn<-NULL
sdd$wave<-NULL
sdd$round<-NULL
sdd$position<-NULL
sdd$positin1<-NULL
sdd$order<-NULL
sdd$partner<-NULL
sdd$pid<-NULL
sdd$undergra<-NULL
sdd$mn_sat<-NULL
sdd$tuition<-NULL
sdd$zipcode<-NULL
sdd$career_c<-NULL
sdd$career<-NULL
sdd$expnum<-NULL
sdd$from<-NULL
sdd$field<-NULL

sdd <- sdd[,c(1:28,49)]

sdd$race<-NULL
sdd$race_o<-NULL

sdd<-sdd[,c(1:5,12:27)]

sdd$like_o<-NULL
sdd$prob_o<-NULL
sdd$met_o<-NULL
sdd$income<-NULL

sdd$match1<-sdd$match
sdd$match<-NULL
sdd$match<-sdd$match1
sdd$match1<-NULL

```

After deleting majority of the columns we have 17 most important variables remaining, which we will use for prediction. 

```{r vars}
pander(colnames(sdd))
```

See the first few rows of the data set in the following table:

```{r head}
sdd <- na.omit(sdd)
```

As a last step I will omit rows with NA values and this will left us with 6838 observations and a clean data set.

```{r omit}
pander(head(sdd))
```

## Description of the final variables

### gender 
Binary variable for gender, (Female=0,Male=1)

```{r gender}
pander(summary(sdd$gender))
qplot(sdd$gender, geom="histogram")
```

### match 
Binary variable which indicates if there was a match or not (1=yes, 0=no). This is the variable that we are going to predict

```{r match}
pander(summary(sdd$match))
qplot(sdd$match, geom="histogram")
```

###int_corr 
Correlation between participants۪s and partners۪s ratings of interests

```{r int_corr}
pander(summary(sdd$int_corr))
qplot(sdd$int_corr, geom="histogram")
```

### samerace 
Participant and the partner were the same race. (1= yes, 0=no)

```{r samerace}
pander(summary(sdd$samerace))
qplot(sdd$samerace, geom="histogram")
```

### age_o 
age of partner in years

```{r age_o}
pander(summary(sdd$age_o))
qplot(sdd$age_o, geom="histogram", binwidth=1)
```

### dec_o 
Decision of partner the night of event (1=yes, 0=no)

```{r dec_o}
pander(summary(sdd$dec_o))
qplot(sdd$dec_o, geom="histogram")
```


### attr_o, sinc_o, intel_o, fun_o, amb_o, shar_o
Rating by partner the night of the event, for all 6 attributes (Attractiveness, Sincerity, Intelligence, Fun, Ambition, and Shared Interests)

```{r attr}
pander(summary(sdd$attr_o))
pander(summary(sdd$sinc_o))
pander(summary(sdd$intel_o))
pander(summary(sdd$fun_o))
pander(summary(sdd$amb_o))
pander(summary(sdd$shar_o))
```

### age 
Age of the person in years
```{r age}
pander(summary(sdd$age))
qplot(sdd$age, geom="histogram", binwidth=1)
```

### field_cd 
Field of study coded:

1. Law  
2. Math
3. Social Science, Psychologist 
4. Medical Science, Pharmaceuticals, and Bio Tech 
5. Engineering  
6. English/Creative Writing/ Journalism 
7. History/Religion/Philosophy 
8. Business/Econ/Finance 
9. Education, Academia 
10. Biological Sciences/Chemistry/Physics
11. Social Work 
12. Undergrad/undecided 
13. Political Science/International Affairs 
14. Film
15. Fine Arts/Arts Administration
16. Languages
17. Architecture
18. Other

```{r field_cd}
pander(summary(sdd$field_cd))
qplot(sdd$field_cd, geom="histogram", binwidth=1)
```

### imprace
How important is it to the person (on a scale of 1-10) that the partner you date be of the same racial/ethnic background?

```{r imprace}
pander(summary(sdd$imprace))
qplot(sdd$imprace, geom="histogram", binwidth=1)
```


### imprelig
How important is it to the person (on a scale of 1-10) that the partner you date be of the same religious background?

```{r imprelig}
pander(summary(sdd$imprelig))
qplot(sdd$imprelig, geom="histogram", binwidth=1)
```

### exphappy
Overall, on a scale of 1-10, how happy the person expect to be with the people she/he meet 
during the speed-dating event?

```{r exphappy}
pander(summary(sdd$exphappy))
qplot(sdd$exphappy, geom="histogram", binwidth=1)

```

## Few regressions, checking correlations between variables based on survey data

### Exhappy - on age
Older persons expects to be more happy with the partner met on the speed date 
```{r eona, warning=FALSE}
ggplot(data = sdd, aes(x=age, y=exphappy)) +
  geom_point(size=1.5, colour="orange") +
  geom_smooth(method="lm", colour="darkgreen") 
```

### Race importance - on age
Importance of the partner to be the same race is becoming less and less with age
```{r ageonimprace, warning=FALSE}
ggplot(data = sdd, aes(x=age, y=imprace)) +
  geom_point(size=1.5, colour="orange") +
  geom_smooth(method="lm", colour="darkgreen") 
```

### Race importance - on religion importance
For persons where the it is more important that the partner is of same religious background it is more important that they are of same racial background
```{r raceonreligion, warning=FALSE}
ggplot(data = sdd, aes(x=imprelig, y=imprace)) +
  geom_point(size=1.5, colour="orange") +
  geom_smooth(method="lm", colour="darkgreen") 
```

### Attractiveness on intelligence
More intelligent persons are more attractive on average
```{r attronintel, warning=FALSE}
ggplot(data = sdd, aes(x=intel_o, y=attr_o)) +
  geom_point(size=1.5, colour="orange") +
  geom_smooth(method="lm", colour="darkgreen") 
```

### Ambition on intelligence
More ambitious persons are more intelligent on average
```{r ambonintel, warning=FALSE}
ggplot(data = sdd, aes(x=amb_o, y=intel_o)) +
  geom_point(size=1.5, colour="orange") +
  geom_smooth(method="lm", colour="darkgreen") 
```

## Creating train and test datasets
For modelling I will create Train and Test data sets. The data sets will be created with randomly choosing observations in the following breakdown:

* Train data set - 50% of data, used for training the models
* Test data set - 50% of data, used for model evaluation

```{r traintest, warning=FALSE}
sdd <- data.table(sdd)
sdd[, rnd := runif(6838)]
setorder(sdd, rnd)
sdd
sdd[, rnd := NULL]

sdd <- na.omit(sdd)

train <- sdd[1:3419]
test  <- sdd[3420:6838]

```

## Modelling

During the modelling I will try to predict if there will be a match or not based on 16 variables.
First I will start with 2 simple models in R:

* K-nearest neighbors - where K=10,15,30
* Decision tree

Then I will predict matches with H2o using the following algorithms:

* Random Forest - with different setups
* GBM - with different setups

### KNN

#### 10NN
```{r 10nn, warning=FALSE}

for (v in c('gender', 'match', 'samerace', 'dec_o', 'field_cd')) {
    set(sdd, j = v, value = as.factor(sdd[, get(v)]))
}

res <- knn(train[, 1:16, with = FALSE], test[, 1:16, with = FALSE], train$match, k = 10)
pander(table(test$match, res))
auc(res,test$match)
plot.roc(res, test$match, main="10NN ROC", percent=TRUE, ci=TRUE, print.auc=TRUE)

```

#### 15NN
```{r 15nn, warning=FALSE}
res2 <- knn(train[, 1:16, with = FALSE], test[, 1:16, with = FALSE], train$match, k = 15)
pander(table(test$match, res2))
auc(res2,test$match)
plot.roc(res2, test$match, main="15NN ROC", percent=TRUE, ci=TRUE, print.auc=TRUE)

```

#### 30NN
```{r 30nn, warning=FALSE}
res3 <- knn(train[, 1:16, with = FALSE], test[, 1:16, with = FALSE], train$match, k = 30)
pander(table(test$match, res3))
auc(res3,test$match)
plot.roc(res3, test$match, main="30NN ROC", percent=TRUE, ci=TRUE, print.auc=TRUE)

```

### Decision Tree

#### DT1
First decision tree calculated with dec_o variable. This variable shows the decision of the partner so it is logical that it will be an important predictor 
```{r dt, warning=FALSE}

for (v in c('gender', 'match', 'samerace', 'dec_o', 'field_cd')) {
    set(train, j = v, value = as.factor(train[, get(v)]))
}

for (v in c('gender', 'match', 'samerace', 'dec_o', 'field_cd')) {
    set(test, j = v, value = as.factor(test[, get(v)]))
}

ct <- rpart(match ~ ., data = train)
summary(ct)


plot(as.party(ct))

res4 <- predict(ct, newdata = test, type = 'class')

pander(table(test$match, res4))
auc(res4,as.numeric(test$match))
plot.roc(res4,as.numeric(test$match), main="DT1", percent=TRUE, ci=TRUE, print.auc=TRUE)

```
 
#### DT2
Second tree was built without dec_o variable to see which other variables are important if we don't know the decision of the partner
```{r dt2, warning=FALSE, eval=FALSE}

train1<-data.table(train)
train1$dec_o<-NULL

test1<-data.table(test)
test1$dec_o<-NULL


ct1 <- rpart(match ~ ., data = train1)
summary(ct1)


plot(as.party(ct1))

res5 <- predict(ct1, newdata = test1, type = 'class')

pander(table(test1$match, res5))
auc(res5,as.numeric(test1$match))
plot.roc(res5, as.numeric(test1$match), main="DT2", percent=TRUE, ci=TRUE, print.auc=TRUE)

```

## H2O setup for modelling

I used R and H2O for modelling purposes. Before start of the H2O modelling I need to setup the H2O environment as well as to convert some of the variables to factors.

```{r h2osetup, warning=FALSE}
library(h2o)
localH2O = h2o.init()

h_train <- as.h2o(train)
h_test <- as.h2o(test)
```

### Random Forest

#### RF1
Setup is: Number of trees=500 and Maximal depth = 10
```{r rf1, warning=FALSE, message=FALSE}
{
  res_rf1 <- h2o.randomForest(x = c(1:16), 
                            y = "match", 
                            training_frame = h_train, 
                            mtries = -1, 
                            ntrees = 500, 
                            max_depth = 10, 
                            nbins = 20)
}

res_rf1

h2o.auc(res_rf1)

h2o.auc(h2o.performance(res_rf1, h_test))
```

#### RF2
Setup is: Number of trees=500 and Maximal depth = 15
```{r rf2, warning=FALSE, message=FALSE}
{
  res_rf2 <- h2o.randomForest(x = c(1:16), 
                            y = "match", 
                            training_frame = h_train, 
                            mtries = -1, 
                            ntrees = 500, 
                            max_depth = 15, 
                            nbins = 20)
}

res_rf2

h2o.auc(res_rf2)

h2o.auc(h2o.performance(res_rf2, h_test))
```

#### RF3
Setup is: Number of trees=1000 and Maximal depth = 15
```{r rf23, warning=FALSE, message=FALSE}
{
  res_rf3 <- h2o.randomForest(x = c(1:16), 
                            y = "match", 
                            training_frame = h_train, 
                            mtries = -1, 
                            ntrees = 1000, 
                            max_depth = 15, 
                            nbins = 20)
}

res_rf3

h2o.auc(res_rf3)

h2o.auc(h2o.performance(res_rf3, h_test))
```

### GBM

#### GBM1
Setup is: Number of trees=100 and Maximal depth = 10
```{r gbm1, warning=FALSE, message=FALSE}
{
  res_gbm1 <- h2o.gbm(x = c(1:16), 
                    y = "match", 
                    training_frame = h_train, 
                    ntrees = 100, 
                    max_depth = 10, 
                    learn_rate = 0.01) 
}

res_gbm1

h2o.auc(res_gbm1)

h2o.auc(h2o.performance(res_gbm1, h_test))
```

#### GBM2
Setup is: Number of trees=500 and Maximal depth = 15
```{r gmb2, warning=FALSE, message=FALSE}
{
  res_gbm2 <- h2o.gbm(x = c(1:16), 
                    y = "match", 
                    training_frame = h_train, 
                    ntrees = 500, 
                    max_depth = 15, 
                    learn_rate = 0.01) 
}

res_gbm2

h2o.auc(res_gbm2)

h2o.auc(h2o.performance(res_gbm2, h_test))
```

#### GBM3
Setup is: Number of trees=1000 and Maximal depth = 15
```{r gmb3, warning=FALSE, message=FALSE}
{
  res_gbm3 <- h2o.gbm(x = c(1:16), 
                    y = "match", 
                    training_frame = h_train, 
                    ntrees = 1000, 
                    max_depth = 15, 
                    learn_rate = 0.01) 
}

res_gbm3
h2o.auc(res_gbm3)
h2o.auc(h2o.performance(res_gbm3, h_test))
```

## Model analysis

* With KNN:
    + 10NN: AUC = 0.6734
    + 15NN: AUC = 0.7068
    + 30NN: AUC = 0.7522

Among KNN the best performing model is where K = 30. It is the model with highest complexity among KNN models tested. Computing time for all of the models was pretty fast.

* With DT: AUC = 0.7042

From the DT we can see that the dec_o predictor was the most important one which is logical since this variable indicates the decision of the partner. The second two important predictors where if they share lots of interests and if the person was funny, which is also logical if take into account that this was a speed date experiment where the people had only 4 minutes to get to know each other and to decide.

```{r varimp, warning=FALSE}
pander(varImp(ct))
```

* With RF:
    + RF1: AUC = 0.8928632
    + RF2: AUC = 0.8947308
    + RF3: AUC = 0.8942053

The best performing RF model was the one with 500 trees and maximal depth of 15 (RF2). This was the middle complexity model and it had the middle computation time amongst the random forest models. We can see from the AUC that the RF3 model is already over fitted a bit compared to RF2. The computation time of all RF models was more than in case of decision tree and KNN models.

* With GBM:
    + GBM1: AUC = 0.8822697
    + GBM2: AUC = 0.8888288
    + GBM3: AUC = 0.8884344

The best performing GBM model was the one with 500 trees and maximal depth of 15 (GBM2). The complexity and computation time was growing from first model to third. The third model has weaker performance from the second so it is over fitted a bit. The GBM models had the highest computation time from all of the models.

## Model selection

The best performing model from the models above was the RF2 random forest model with 500 trees and 15 depth. 
